---
title: "Hour of power topics: missing data"
output: html_document
---

First of all it is important to determine whether [data is missing at random or not](https://en.wikipedia.org/wiki/Missing_data#Missing_completely_at_random). If there is no structure in the missing data, one could take a number of decisions about how to impute it. Here I will focus in 

Let's produce an artificial dataset:

```{r,cache=F,echo=T}
set.seed(300)
like.maths<-sample(c("Y","N"),500,replace=T)
sex<-sample(c("M","F"),500,replace=T)
height<-sapply(sex,function(x) ifelse(x=="M",rnorm(1,1.80,0.2),rnorm(1,1.60,0.2)))
education<-sample(c(1:5),500,replace=T)
iq<-rnorm(500,100,20)
test.data<-data.frame(sex=sex,height=height,education=education,iq=iq,like.maths=like.maths,stringsAsFactors=F)
grade<-apply(test.data,1,function(x) 10*plogis(rnorm(1,0,2)+ifelse(x["like.maths"]=="Y",0.3,-0.3)+0.2*(as.numeric(x["education"])-3)+(as.numeric(x["iq"])-100)/10)) 
test.data$iq[sample(c(1:500),100)]<-NA
test.data$height[sample(c(1:500),100)]<-NA
test.data$sex[sample(c(1:500),100)]<-NA
test.data$education[sample(c(1:500),100)]<-NA
test.data$grade<-grade
print(str(test.data))
```

Let's check the distribution of grades:

```{r,echo=T,cache=F}
print(hist(test.data$grade))
```

For instance, let's check the relation between grade and iq
```{r,echo=T,cache=F}
plot(grade~iq,data=test.data)
```

We can get a correlation values from this, for instance
```{r,echo=T}
correlation<-cor(test.data$grade,test.data$iq,use="complete.obs")
print(correlation)
```

The question is then: how much would variation would have been if we had observed the missing data?

A very reasonable assumption is that, other things being equal, the distribution of the values in the missing cells corresponds to. Let's define a small function that does this:

```{r,echo=T,cache=T}
boot.md<-function(x) {
x[is.na(x)]<-sample(x[!is.na(x)],length(x[is.na(x)]))  
return(x)}
```

Instead of sampling one time, we should produce multiple imputations and 

```{r,echo=T}
icor<-rep(0,500)
for(i in c(1:500)) {
iiq<-boot.md(test.data$iq)
icor[i]<-cor(iiq,test.data$grade,use="complete.obs")}
```

Now we can see the values of these imputed datasets:

```{r,echo=T}
hist(icor)
```

Naturally, the correlations are smaller because we introduced unstructered data!